Quantifying the Importance of Change for Understanding Differences in Personal Culture

### 2024-03-14 manuscript/manuscript.Rmd
kkedits314

---
title: "Quantifying The Importance of Change for Understanding Differences in Personal Culture"
author: 
- 
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
geometry: margin=1in 
fontsize: 12pt
header-includes:
  \usepackage{fontspec}
  \setmainfont{Times New Roman}
  \setlength{\parindent}{3em}
  \usepackage{caption}
  \usepackage{setspace}\doublespacing
  \usepackage{endnotes}
  \renewcommand{\enotesize}{\normalsize}
  \let\footnote=\endnote
  \pagenumbering{gobble}
bibliography: kileybib.bib
csl: asa.csl
abstract: "When researchers want to understand differences in personal culture---a person's attitudes, beliefs, values, and practices--how much attention should they pay to adult experience? Recent work has reached substantially different conclusions on this question. We argue that this disagreement is an unintended consequence of the \"tournament of models\" approach researchers have used, which focuses on whether people change and not how much they change. To advance the theoretical debate, we refocus attention on the relative importance of personal change over time for explaining differences between people. We introduce a new measure for quantifying the proportion of systematic variance in panel data attributable to intrapersonal change. Applying this measure to 609 items from seven surveys in five countries, we find that although intrapersonal change is common, it is generally small in magnitude. As an extension of the theoretical debate, we demonstrate that this measure provides new insights when comparing social groups, showing that intrapersonal change is less common among U.S. college graduates than among those without a college degree. Our findings provide a new perspective on several important theoretical debates, as well as a tool to address new questions."
---

@@

@@

Does personal culture---a person's attitudes, beliefs, values, and practices---change over the life course, or is it largely fixed by adulthood? This question underlies an important contemporary debate in sociology [@lizardo2017; @kiley2020; @lersch2023] and has deep roots in seemingly contradictory theoretical perspectives. For example, pragmatist theories of action claim that changes in social environments cause people to adapt their views and make new cultural meanings [@swidler2001; @gross2009], while Bourdieusian practice theories argue that the "past conditions of production" leave a mark on people's personal culture that lasts throughout their lives [@bourdieu1990]. Models of social influence assume that people adapt their culture in the face of new information [@goldberg2018], while the emphasis on cohort effects in models of aggregate social and cultural change requires them to be open to change while young but become fairly resistant to it as they age [@ryder1965]. Finally, life course theories posit important changes over time as people advance through important transitions in their lives [@elder2003; @bardi2009].

Because all of these perspectives have some empirical support, the theoretical debate is not about whether the processes they posit exist at all, but about their relative contribution to explaining the cultural differences we see in the world. Nevertheless, researchers have struggled to reach a consensus on the importance of cultural change during adulthood. Over the span of a few years, most apparent changes in adults' survey responses appear to be transitory, with little evidence of persistent change [@kiley2020; @vaisey2021]. This suggests that change during adulthood may not be a major factor in explaining contemporary cultural divides. By contrast, when we consider a longer time horizon, there is evidence that adults make at least some persistent changes [@lersch2023].

These seemingly inconsistent findings are partly due to the fact that researchers have taken a "tournament of models" approach to adjudicate different theoretical perspectives [@lersch2023: p. 228]. That is, researchers have relied on model selection criteria---primarily the Bayesian Information Criterion [@raftery1995]---to sort variables into the "no change" or "any change" piles. The size of these piles then serves as the primary evidence for the truth of the theory [@kiley2020; @vaisey2021; @lersch2023].

We argue, however, that asking *whether* people change on an item is an unintended source of confusion. Being able to detect *any* change on an item depends not only on sample populations, measurement, statistical methods, and power, but also on definitions of what counts as "change." These differences can limit the possibility of agreement on the assignment of a variable to the right pile. Moreover, asking whether there is evidence of change reduces to, "Is there *any* evidence of change in this item?" But such binary questions are ill-suited to address the core of the theoretical debate: determining *how much* the process of intrapersonal change contributes to cultural differences.

Of course neither pre-adult socialization nor the contemporary social context alone can explain the full range of cultural differences. The relevant question is about the *relative* contributions of these processes. In the current debate, for a single construct (e.g., support for gay rights), model comparison approaches can only help decide *whether* intrapersonal change is happening, not about how much. This also prevents investigating under what conditions, across which groups, and in what domains we might see differences in the amount of intrapersonal change. These questions are about *degree*, not existence. Taking this debate forward therefore requires precise measurement rather than declarations of victory for a particular perspective.

In this paper, we move beyond the "tournament of models" approach and introduce a new method for quantifying the relative contributions of interpersonal differences and intrapersonal change for a single item. We use 609 items from seven panel surveys from five countries to quantify the proportion of variance that is attributable to systematic intrapersonal change over the study period rather than to stable interpersonal differences.

We observe similar results across all datasets despite their varied social contexts and duration. In general, we find that intrapersonal change accounts for a small amount of variance in personal culture survey items. Of course, any question about whether a given amount of intrapersonal change is "meaningful" or "important" is substantive and theoretical. Nevertheless, on many measures (some observed for over a decade), we do not see enough change to believe that intrapersonal change processes play a substantial role in explaining the differences we observe between adults. Notable, interesting, and important exceptions exist, but they are departures from the overall pattern.

In addition to advancing a theoretical debate, our method can also help address specific substantive questions. To illustrate how it might be used by a researcher with a specific topical interest, we investigate differences in the relative importance of intrapersonal change for people with and without a college degree. We find a smaller amount of systematic change among college-educated respondents, suggesting that college crystallizes one's personal culture rather than fostering an openness to new information. This offers a new empirical basis to theorize about the role of education in influencing personal culture.

Taken together, our goal in this paper is to advance---and, hopefully, transcend---debates about *whether* adults change. People do change, at least a little, on most things. Our method can help quantify exactly *how much*.

@@

Empirically comparing which of these models better fit a broad range of questions from the General Social Survey's rotating panels, Kiley and Vaisey [-@kiley2020] found limited evidence of durable change. While the majority of items did prefer the AUM, the amount of durable change detected on these items was small. A substantial minority of questions (39 percent) favored the SDM, meaning they were more consistent with zero durable change. Questions with more evidence of durable change included salient issues like gay marriage and questions tapping "public" statements or behaviors such as partisan identification and religious service attendance. There was also more evidence of durable change among early adults (people ages 18-30) than among the rest of the population. Overall, the researchers concluded that "results ultimately suggest that real, persistent attitude change is an uncommon phenomenon among adults'' [@kiley2020: p. 500; see also @vaisey2021]. This lack of durable change is consistent with other recent findings that cohort replacement plays a somewhat larger role than period effects in explaining differences in personal culture [@vaisey2016].

@@

Drawing on these findings, Lersch [-@lersch2023] challenged the SDM and AUM as a "needless dichotomy," proposing the "Life Course Adaption Model" (LCAM) as an alternative. This model draws on the life course perspective to model personal culture as a different linear trajectory over the duration of a panel for each respondent. In doing this, Lersch rectified two shortcomings of the AUM and SDM. First, the AUM, as described by Kiley and Vaisey, posits that changes follow a Markov process where responses at time $t$ are a function of responses at time $t-1$ but not earlier time points. However, earlier life experiences can (directly and indirectly) mold personal culture when transitioning to new social roles or into new environments, even if their initial impact is delayed. For example, childhood events might influence views on family structures later when individuals form their own families. Therefore, the LCAM considers influences from earlier than just $t-1$ on responses at time $t$.

Second, Kiley and Vaisey's analysis of the AUM and SDM is based on three-wave panel data over four years, which might not be intensive or extensive enough to adjudicate the two models. Lersch evaluates the LCAM against the AUM and SDM using panel data spanning a wider duration (from 3 to 36 years) and more waves (3 to 18), offering a better chance to observe durable change.

When Lersch compared the LCAM to the AUM and SDM on data from five countries, 297 of the 428 questions he analyzed preferred the LCAM, suggesting that we observe some linear change for adults on most items. The SDM was preferred on 112 items, and the rest did not yield conclusive results. No questions favored the AUM. He concludes that "new experiences over the life course [...] can persistently move individuals' personal culture in novel directions" [@lersch2023: p. 243-244].

Despite differences in how these researchers interpret their findings, the empirical results are not far apart. Kiley and Vaisey [-@kiley2020] found that the majority of items they tested favored the AUM, meaning there was evidence of durable change among adults, a pattern consistent with Lersch's results. Also Lersch [-@lersch2023] found evidence of durable change on most items, but these changes were small. On average, people changed only about .07 standard deviations over 10 years. This is consistent with Kiley and Vaisey's finding that durable change is often small. Moreover, a quarter of the items Lersch studied still preferred the SDM, suggesting that even with different assumptions about change and more extensive data, many questions still are consistent with a model that allows zero intrapersonal change.

@@

To this point, the debate has been about whether people *ever* change. As we argued above, this is rarely the right question since the answer is almost always "yes." Over sufficient time and in a large enough sample, researchers will likely observe some evidence for durable individual change. A lack of evidence for change may be due to a poor survey question, a low resolution in response options, or that the question simply was not asked for long enough or to enough people. Conversely, finding evidence for some change tells us only that it was "not zero." From that alone, we learn little about how much intrapersonal change has happened in a population.

A more theoretically productive approach begins with a model assuming that during an observed time period people might remain stable, might change a little, or might undergo significant shifts in their personal culture. Lersch's LCAM does this by modeling each individual as following a personal linear trajectory. But rather than debating whether this model fits a particular data set better than a model that assumes that people never change, it would be better to quantify these individual changes and compare them to the stable differences that exist between people. In other words, the focus should shift from asking, "do people change?" to asking, "what are the relative contributions of change and pre-existing differences for explaining variance in personal culture?"

We propose an approach---which we formalize below---that offers such a quantification. Our measure relates two variance components that each reflect a combination of relevant theoretical processes. The first component, stable interpersonal differences, reflects the accumulated experiences of people prior to entering the panel survey. Lersch [-@lersch2023: 24] calls this "early imprinting." While commonly associated with experiences during a formative period that result in settled dispositions, this variation could also reflect experiences that happened at any time as long as they predate the panel and consistently affect subsequent responses. For instance, for those people entering the panel post-retirement, this "imprinting" might reflect this pivotal life transition. Consequently, this component also reflects variation in individuals' social roles or statuses at the start of the panel that were important in shaping their dispositions.

The second component, the amount of intrapersonal linear change people make during the panel, captures durable changes in personal culture over time. This captures the set of processes collectively called "persistent change" or "adaption" by Lersch and "active updating" by Kiley and Vaisey. Lersch attributes these changes to social triggers such as moving into a new environment or adopting new social roles (although he does not measure these directly). They might also reflect the diffusion of new cultural forms across social networks, cues from political elites or otherwise culturally influential leaders, the emergence of issues in politics or culture, or large-scale social shifts.

@@

Furthermore, these components are not necessarily useful in isolation or without context. What is important for the theoretical debate is the relative contribution of these two components in explaining cultural differences in a population. In other words, when we look at a population of adults over some time frame, is there enough intrapersonal change relative to interpersonal difference to view it as an important process for why people differ from each other? Only through quantifying the two components we can get closer to a true answer to this question.

@@

@@

## Life Course Adaption Model

The estimation procedure begins with Lersch's [-@lersch2023] Life Course Adaption Model, which formalizes survey responses at time $t$ as a function of individual-level random intercepts and slopes for survey age. In some contexts, this model is called a mixed-effects growth curve model. This model assumes a set of propositions about change that reflect the theoretical debate to this point. First, consistent with the settled disposition model, it assumes that people start the survey with cultural differences, modeled as random intercepts for each respondent. Second, it assumes that people change over time, taking the form of random slopes for each respondent as a linear function of time. Third, it assumes that people deviate around this baseline randomly over time, reflecting "fluctuation" or short-term non-persistent change. Formally, this can be written as

@@

As a first step in our analysis, we fit this LCAM to each of the 609 measures of personal culture outlined above. We then use these models to derive two measures of the components we outlined above: pre-existing interpersonal differences and intrapersonal change over time.

The third component, which is the proportion of variance attributable to measurement error or transient fluctuations, is what's left, such that these three numbers sum to 1. As we discussed above, this component is not of core interest here.

As a second step in our analysis, we use linear regression to conduct a meta-analysis, modeling $\omega$ as a function of features of the question and panel. This allows us to explore how these features are associated with the degree of intrapersonal change. These covariates include the number of response options (measured as 2, 3-5, and more than 5), the specific survey, the log of the number of participants, the number of waves observed, the time period (i.e., decade) the question was first asked, and the total duration of time the question was asked.

@@

The objective of this paper is to intervene in the recent debate on whether people change their personal culture -- their attitudes, beliefs, values, and practices -- as they move through their adult life. Instead of falling back on a "tournament of models" approach [@lersch2023: p. 228] to proclaim victory for a particular answer to this question, our goal is to refocus the debate to the theoretically more productive question of the relative importance of intrapersonal change for explaining differences in the personal culture. We propose a new approach that quantifies the amount of observed systematic variance that is attributable to either interpersonal differences at baseline or intrapersonal change over the duration of a panel.

@@

More than adjudicating past positions in this debate, the approach we presented in this paper will enable scholars to explore more precisely the relative contributions of change and stability in explaining variation in personal culture, as well as how these contributions might differ across populations. To illustrate this, we showed how the amount of cultural difference explained by intrapersonal change varies substantially by survey item but also by individuals' characteristics such as their education. While age and other factors related to college completion might confound this pattern, it suggests that college completion may crystallize personal culture to an extent that renders later adult experiences on political dispositions less influential.

@@

Although we believe it is a major advance on previous work, the approach we outline here has important limitations. We allocate all systematic variance to one of two sets of theoretical processes: intrapersonal change and interpersonal differences at baseline. Our approach does not quantify the proportion of people who "change," nor can we be sure that the amount of interpersonal change we detect is driven by many people making small changes or a few people making large changes.

Our results and interpretations also hinge on how we have defined change. As does the LCAM, our approach treats cultural trajectories as varying linear slopes for each respondent, thus assuming that change is a linear function of time. This assumption simplifies reality in which change likely also takes non-linear and discontinuous forms. People might jump from one "stable" disposition to another or experience a "turning point" in their lives that upsets an otherwise stable trajectory. In parts, our assumption of linearity is a limitation of the data, as most questions are only observed for three to six waves. Panels with more waves might allow researchers to loosen this assumption to test alternative, more flexible models of change.

Similarly, our approach assumes that durable change is unidirectional. This is a sensible assumption on short panels where classifying change that lasts less than two years as durable seems unreasonable. Practically, it means that the variance produced by people making durable changes in their cultural dispositions to then return to a previous state later in life is classified as residual variance, rather than intrapersonal change. Longer panels and more flexible definitions of "change" might allow us to account for such trajectories.

@@

Despite these limitations, we believe our method and findings have important implications for social science research. Sociologists interested in understanding cultural differences have largely asked about the *existence* of cultural change in adults [e.g., @kiley2020; @vaisey2021; @lersch2023]. Yet this approach has inadvertently limited the debate. In any population, *some* degree of adult cultural change is inevitable. Although ideal types like the settled dispositions and active updating models are useful, no theoretical perspective would expect either early life socialization or adult intrapersonal change to be the sole source of one's personal culture. Our results reinforce this point, showing the relevance of both factors and allowing their precise quantification.

This and other recent findings [@stewart2023; @quinn2023] suggest that it is theoretically more productive to measure the relative importance of these two components in concrete, substantive cases. Drawing a unified conclusion from survey items measuring various cultural forms on different scales and across different time frames is challenging. Nevertheless, the general pattern suggests that, for most items, intrapersonal change in adulthood is not the primary reason for the differences we see between people in the world.

@@

We believe the approach we outlined here can push past the "needless dichotomy" implicit in the question of whether people change or not [@lersch2023]. Characterizing questions as displaying change or not can only take researchers so far, but the question of whether some questions demonstrate more change than others, or whether some groups are characterized by more stability than others, has the potential to weigh in on a broader range of theoretical debates. We hope researchers find our approach useful as they investigate these questions.

@@

>>>>

---
title: "Quantifying the Importance of Change for Understanding Differences in Personal Culture"
author: 
  - Achim Edelmann^[médialab, Science Po]
  - Kevin Kiley^[Department of Sociology and Anthropology, North Carolina State University]
  - Turgut Keskintürk^[Department of Sociology, Duke University]
  - Isabella Bouklas^[Department of Sociology, Duke University]
  - Stephen Vaisey^[Department of Sociology, Duke University]
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
geometry: margin=1in 
fontsize: 12pt
header-includes:
  \usepackage{fontspec}
  \setmainfont{Times New Roman}
  \setlength{\parindent}{3em}
  \usepackage{caption}
  \usepackage{setspace}
  \usepackage{endnotes}
  \renewcommand{\enotesize}{\normalsize}
  \let\footnote=\endnote
bibliography: kileybib.bib
csl: asa.csl
abstract: "Work in the sociology of culture has converged on a general conclusion that while people sometimes make substantial shifts in their personal culture over time, most elements of personal culture are characterized by low rates of persistent change during adulthood. Recent developments have begun to quantify change over time and initial differences, but it is unclear how to use these components to understand the relative importance of change, especially when change is modeled non-linearly. To advance this debate, we introduce a measure for quantifying the proportion of systematic variance in panel data attributable to intrapersonal change. Applying this measure to 609 items from seven surveys in five countries, we find that although intrapersonal change is common, it does not explain a large share of systematic variance. As an extension of the theoretical debate, we demonstrate that this measure provides new insights when comparing social groups, showing that intrapersonal change is less common among U.S. college graduates than among those without a college degree. Our findings provide a new perspective on several important theoretical debates, as well as a tool to address new questions."
---

@@

\doublespacing

@@

An important contemporary debate in sociology centers on the importance of intrapersonal change for understanding differences in "personal culture" -- attitudes, beliefs, values, and practices [@lizardo2017; @kiley2020; @lersch2023; @underwood2021; @vaisey2016]. This question has deep roots in seemingly contradictory theoretical perspectives. For example, pragmatist theories of action claim that changes in social environments cause people to adapt their views and make new cultural meanings [@swidler2001; @gross2009], while Bourdieusian practice theories argue that the "past conditions of production" leave a mark on people's personal culture that lasts throughout their lives [@bourdieu1990]. Models of social influence assume that people adapt their culture in the face of new information [@goldberg2018], while the emphasis on cohort effects in models of aggregate social and cultural change requires them to be open to change while young but become fairly resistant to it as they age [@ryder1965]. Finally, life course theories posit important changes over time as people advance through important transitions in their lives, but that certain life stages have more lasting influence than others [@elder2003; @bardi2009].

Recent work on this topic has converged on a general conclusion that people do sometimes make substantial shifts in their personal culture over time but that most elements of personal culture are characterized by low rates of persistent change during adulthood [@kiley2020; @lersch2023]. Until recently, this work has adopted a "tournament of models" approach to adjudicating different theoretical perspectives [@lersch2023: p. 228]. In this framework, researchers formalized different models of change over time and relied on model selection criteria---primarily the Bayesian Information Criterion [@raftery1995]---to classify different questions as demonstrating either "no change" or "any change." The size of these piles then serves as the primary evidence for the truth of different arguments [@kiley2020; @vaisey2021].

However, as consensus in the overall picture emerges, researchers are now turning toward understanding the conditions that produce persistent change in personal culture, including age, education, and life-course transitions; whether certain kinds of people are more likely to make persistent changes in personal culture; and whether certain kinds of personal culture are more likely to be characterized by persistent change or stability. The tournament of models approach, useful for the broad question of whether change happens, is ill-suited to addressing these more theoretically grounded questions.

In a recent paper, Lersch [-@lersch2023] takes an important step forward by devising a Life Course Adaption Model (LCAM), a mixed-effects growth curve model that quantifies various components of cultural difference: early imprinting, persistent change, biographical experience, and fluctuation. In quantifying these aspects of cultural difference, the LCAM lets researchers address a broader range of questions than the single question of whether a measure is characterized by stability or change. However, because the LCAM produces so many different measures, it is not obvious how to use these components to understand the relative importance of change, espeically when change is measured in non-linear ways.

In this paper, we build on the LCAM by outlining a single metric that combines diverse components of the LCAM to quantify the relative importance of intrapersonal change and interpersonal difference for explaining variation in a population for a variable over time. Our approach isolates the "systematic" components of variance (i.e., it removes random change and measurement error), to provide a single succinct summary of how much systematic variation can reasonably be attributed to people making persistent changes over time. We demonstrate the utility of this quantity in two ways.

First, we use 609 items from seven panel surveys from five countries to quantify the proportion of variance that is attributable to systematic intrapersonal change over the study period rather than to interpersonal differences at baseline. This allows us to generate an overall picture of how much of the total systematic variation in personal culture change tends to explain. These findings largely echo previous findings but enable clearer statements about the relative importance of change across questions, rather than binary claims about whether change "happens" or specific quantification of how much change occurs.

Second, to illustrate how this quantity might be used by a researcher with a specific topical interest, we investigate differences in the relative importance of intrapersonal change for people with and without a college degree. We find that change explains a smaller proportion of systematic variance among college-educated respondents, suggesting that college crystallizes one's personal culture rather than fostering an openness to new information. This offers a new empirical basis to theorize about the role of education in influencing personal culture.

@@

Empirically comparing which of these models better fit a broad range of questions from the General Social Survey's rotating panels, Kiley and Vaisey [-@kiley2020] found limited evidence of durable change. While the majority of items did prefer the AUM, the amount of durable change detected on these items was small. A substantial minority of questions (39 percent) favored the SDM, meaning they were more consistent with zero durable change. Questions with more evidence of durable change included salient issues like gay marriage and questions tapping "public" statements or behaviors such as partisan identification and religious service attendance. There was also more evidence of durable change among early adults (people ages 18-30) than among the rest of the population. Overall, the researchers concluded that "results ultimately suggest that real, persistent attitude change is an uncommon phenomenon among adults'' [@kiley2020: p. 500; see also @vaisey2021]. This lack of durable change is consistent with other recent findings that cohort replacement plays a larger role than period effects in explaining differences in personal culture [@vaisey2016].

@@

Drawing on these findings, Lersch [-@lersch2023] challenged the SDM and AUM as a "needless dichotomy," proposing the "Life Course Adaption Model" (LCAM) as an alternative. This model draws on the life course perspective to model personal culture as a different linear trajectory[^linear] over the duration of a panel for each respondent. In doing this, Lersch rectified three shortcomings of the AUM and SDM. First, the AUM, as described by Kiley and Vaisey, posits that changes follow a Markov process where responses at time $t$ are a function of responses at time $t-1$ but not earlier time points. However, earlier life experiences can (directly and indirectly) mold personal culture when transitioning to new social roles or into new environments, even if their initial impact is delayed. For example, childhood events might influence views on family structures later when individuals form their own families. Therefore, the LCAM considers influences from earlier than just $t-1$ on responses at time $t$.

[^linear]: In supplemental materials, Lersch models age using splines as well, with results generally showing the same conclusions.

Second, Kiley and Vaisey's conclusions about the AUM and SDM were based on examining three-wave panel data over four years, which might not be intensive or extensive enough to adjudicate the two models. Lersch evaluates the LCAM against the AUM and SDM using panel data spanning a wider duration (from 3 to 36 years) and more waves (3 to 18), offering a better chance to observe durable change.

Third, and most importantly, the LCAM quantifies various components of cultural difference, allowing researchers to ask a more varied set of questions than simply whether people change. For example, researchers can use the LCAM framework to quantify the magnitude of change in a measure of culture over time (i.e., how much do people change on average), the variance in age-based slopes over time (i.e., do people tend to change in similar ways as they age), or how much personal culture tends to fluctuate around these age-based trends (how consistent are responses over time).

When Lersch compared the LCAM to the AUM and SDM on data from five countries, 198 of the 428 questions he analyzed preferred the LCAM and 113 preferred the AUM, suggesting that we observe some change for adults on most items. The SDM was preferred on 93 items, and the rest did not yield conclusive results. He concludes that "new experiences over the life course [...] can persistently move individuals' personal culture in novel directions" [@lersch2023: p. 243-244].

Despite differences in how these researchers interpret their findings, the empirical results are not far apart. Kiley and Vaisey [-@kiley2020] found that the majority of items they tested favored the AUM, meaning there was evidence of persisting change among adults, a pattern consistent with Lersch's results. Also Lersch [-@lersch2023] found evidence of persisting change on most items, but these changes were small. On average, people changed only about .07 standard deviations over 10 years. This is consistent with Kiley and Vaisey's finding that persisting change is often small. Moreover, a quarter of the items Lersch studied still preferred the SDM, suggesting that even with different assumptions about change and more extensive data, many questions still are consistent with a model that allows zero intrapersonal change.

@@

To this point, the debate has broadly centered on drawing conclusions about whether people *ever* change. But this is rarely the right question since the answer is almost always "yes." Over sufficient time and in a large enough sample, researchers will likely observe some evidence for durable change in the population. A lack of evidence for change may be due to a poor survey question, a low resolution in response options, or that the question simply was not asked for long enough or to enough people. Conversely, finding evidence for some change tells us only that it was "not zero." From that alone, we learn little about how much intrapersonal change has happened in a population.

A more theoretically productive approach begins with a model assuming that during an observed time period people might remain stable, might change a little, or might undergo significant shifts in their personal culture. Lersch's LCAM does this by modeling each individual as following a personal linear trajectory. In doing so, the LCAM opens the door to a variety of questions. We assert that among the many things that can be done with the LCAM, it provides the opportunity to shift from asking, "do people change?" to asking, "what are the relative contributions of change and pre-existing differences for explaining variance in personal culture?"

Our approach---which we formalize below---offers such a quantification. Our measure relates two variance components that each reflect a combination of relevant theoretical processes. The first component, stable interpersonal differences, reflects the accumulated experiences of people prior to entering the panel survey. Lersch [-@lersch2023: 24] calls this "early imprinting." While commonly associated with experiences during a formative period that result in settled dispositions, this variation could also reflect experiences that happened at any time as long as they predate the panel and consistently affect subsequent responses. For instance, for those people entering the panel post-retirement, this "imprinting" might reflect this pivotal life transition. Consequently, this component also reflects variation in individuals' social roles or statuses at the start of the panel that were important in shaping their dispositions.

The second component, the amount of intrapersonal change people make during the panel, captures persisting changes in personal culture over time. This component captures the set of processes collectively called "persistent change" or "adaption" by Lersch and "active updating" by Kiley and Vaisey. Lersch attributes these changes to social triggers such as moving into a new environment or adopting new social roles (although he does not measure these directly). They might also reflect the diffusion of new cultural forms across social networks, cues from political elites or otherwise culturally influential leaders, the emergence of issues in politics or culture, or large-scale social shifts.

@@

Furthermore, these components are not necessarily useful in isolation or without context. If absolute change is about .07 standard deviations every 10 years, but the variance of starting points at age 18 is very high, then change over time will not produce significant differences among people, even over a long time frame. If, on the other hand, people tend to start at very similar positions and change at a rate of .07 standard deviations, then that change is very useful for understanding difference. What is important for the theoretical debate is the relative contribution of these two components in explaining cultural differences in a population. In other words, when we look at a population of adults over some time frame, is there enough intrapersonal change relative to interpersonal difference to view it as an important process for why people differ from each other? Only through quantifying the two components we can get closer to a true answer to this question.

@@

## Formalization

The starting point for our quantity is Lersch's [-@lersch2023] Life Course Adaption Model, which formalizes survey responses at time $t$ as a function of individual-level random intercepts and slopes for survey age. In some contexts, this model is called a mixed-effects growth curve model. This model assumes a set of propositions about change that reflect the theoretical debate to this point. First, consistent with the settled disposition model, it assumes that people start the survey with cultural differences, modeled as random intercepts for each respondent. Second, it assumes that people change over time, taking the form of random slopes for each respondent as a linear function of time. Third, it assumes that people deviate around this baseline randomly over time, reflecting "fluctuation" or short-term non-persistent change. Formally, this can be written as

We then use these models to derive two measures of the components we outlined above: pre-existing interpersonal differences and intrapersonal change over time.

The third component, which is the proportion of variance attributable to measurement error or transient fluctuations, is what's left, such that these three numbers sum to 1. As we discussed above, this component is not of core interest here, so we set it aside in our analysis.

## Bootstrapping for Confidence Intervals

[Explain how we do that here]

@@

@@

As a first step in our analysis, we fit this LCAM to each of the 609 measures of personal culture
outlined above. We then derive $\omega$ for each of these items.

As a second step in our analysis, we use linear regression to model $\omega$ as a function of features of the question and panel. This allows us to explore how these features are associated with the degree of intrapersonal change. These covariates include the number of response options (measured as 2, 3-5, and more than 5), the specific survey, the log of the number of participants, the number of waves observed, the time period (i.e., decade) the question was first asked, and the total duration of time the question was asked.

@@

The primary objective of this paper is to introduce a new quantity that can advance debate on whether people change their personal culture -- their attitudes, beliefs, values, and practices -- as they move through their adult life. Instead of falling back on a "tournament of models" approach [@lersch2023: p. 228] to proclaim victory for a particular answer to this question, our goal is to refocus the debate to the theoretically more productive question of the relative importance of intrapersonal change for explaining differences in personal culture. We propose a new approach that quantifies the amount of observed systematic variance that is attributable to either interpersonal differences at baseline or intrapersonal change over the duration of a panel.

@@

More than adjudicating past positions in this debate, the approach we presented in this paper will enable researchers to explore more precisely the relative contributions of change and stability in explaining variation in personal culture, as well as how these contributions might differ across populations. To illustrate this, we showed how the amount of cultural difference explained by intrapersonal change varies substantially by survey item but also by individuals' characteristics such as their education. While age and other factors related to college completion might confound this pattern, it suggests that college completion may crystallize personal culture to an extent that renders later adult experiences on political dispositions less influential.

@@

Although we believe the approach we outline here advances previous work, it has important limitations. We allocate all systematic variance to one of two sets of theoretical processes: intrapersonal change and interpersonal differences at baseline. Our approach does not quantify the proportion of people who "change," nor can we be sure that the amount of interpersonal change we detect is driven by many people making small changes or a few people making large changes.

Our results and interpretations also hinge on how we have defined change. As does the LCAM, our analysis here treats cultural trajectories as varying linear slopes for each respondent, thus assuming that change is a linear function of time. This assumption simplifies reality in which change likely also takes non-linear and discontinuous forms. People might jump from one "stable" disposition to another or experience a "turning point" in their lives that upsets an otherwise stable trajectory. In parts, our assumption of linearity is a limitation of the data, as most questions are only observed for three to six waves. However, our overall approach does not require modeling change as a linear function of age. Change can be modeled in non-linear forms, such as the splines used by Lersch [-@lersch2023], and accordingly incorporated into the calculation of variance explained.

Similarly, our approach assumes that durable change is unidirectional. This is a sensible assumption on short panels where classifying change that lasts less than two years as durable seems unreasonable. Practically, it means that the variance produced by people making durable changes in their cultural dispositions to then return to a previous state later in life is classified as residual variance, rather than intrapersonal change. Longer panels and more flexible definitions of "change" might allow us to account for such trajectories. Again, while we model change in a single way, the general variance decomposition approach we outline accommodates different definitions and functional forms of change.

@@

Despite these limitations, we believe our method and findings have important implications for social science research. Sociologists interested in understanding cultural differences have largely asked about the *existence* of cultural change in adults [e.g., @kiley2020; @vaisey2021; @lersch2023]. In any population, *some* degree of adult cultural change is inevitable. Although ideal types like the settled dispositions and active updating models are useful, no theoretical perspective would expect either early life socialization or adult intrapersonal change to be the sole source of one's personal culture. Our results reinforce this point, showing the relevance of both factors and allowing their precise quantification.

This and other recent findings [@stewart2023; @quinn2023; @quinn2024] suggest that it is theoretically more productive to measure the relative importance of these two components in concrete, substantive cases. Drawing a unified conclusion from survey items measuring various cultural forms on different scales and across different time frames is challenging. Nevertheless, the general pattern suggests that, for most items, intrapersonal change in adulthood is not the primary reason for the differences we see between people in the world.

@@

We believe the approach we outlined here can push past the "needless dichotomy" implicit in the question of whether people change or not. Characterizing questions as displaying change or not can only take researchers so far, but the question of whether some questions demonstrate more change than others, or whether some groups are characterized by more stability than others, has the potential to weigh in on a broader range of theoretical debates. We hope researchers find our approach useful as they investigate these questions.

@@

\singlespace
